---
title: "Project ISyE 7401: Predicting H-Index Using Regression Methods"
author: "Moe Kyaw Thu"
date: "2024-03-12"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: readable
    code_folding: show
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)

library(knitr)
library(kableExtra)
library(readr)
library(ggplot2)
library(dplyr)
library(car)
library(stats)
library(lmtest)
library(gghalfnorm)
library(faraway)
library(MASS)
library(caret)
library(stargazer)
library(quadprog)
library(glmnet)
library(lars)
library(nnGarrote)
library(corrplot)
```

# Introduction

This project investigates the factors associated with a researcher's **H-index** using a dataset on scientific productivity and gender. We explore linear regression models with interaction and quadratic terms, perform model diagnostics, and apply regularization techniques including Ridge regression, Lasso, and the Non-Negative Garrote.

# Data Loading and Cleaning

We load a sample of 5,000 observations from the full dataset (~130 million rows) for illustrative purposes.

```{r load-data}
df <- read.delim("scisci_gender.tsv", nrow = 5000)
```

We remove rows with missing values and subset to the columns of interest:

- **H.index** -- H-index (response variable)
- **Productivity** -- Publication count
- **Average_C10** -- Average 10-year citation count
- **Average_LogC10** -- Log-transformed average 10-year citation count
- **Inference_Sources** -- Number of inference sources
- **Inference_Counts** -- Number of inference counts
- **P.gf.** -- Probability of female gender

```{r clean-data}
df.clean <- na.omit(df)

df.clean.sub <- subset(
  df.clean,
  select = c("H.index", "Productivity", "Average_C10",
             "Average_LogC10", "Inference_Sources",
             "Inference_Counts", "P.gf.")
)

df.clean.sub$Average_C10 <- round(df.clean.sub$Average_C10, 2)
df.clean.sub$Average_LogC10 <- round(df.clean.sub$Average_LogC10, 2)
df.clean.sub$P.gf. <- round(df.clean.sub$P.gf., 2)
```

After cleaning, we retain **`r nrow(df.clean.sub)`** observations.

# Exploratory Data Analysis

## Distribution of Variables

```{r eda-distributions, fig.height=10}
visnorm <- function(dataframe) {
  par(mfrow = c(2, ncol(dataframe) %/% 2 + 1))
  for (col in names(dataframe)) {
    hist(dataframe[[col]],
         main = paste("Histogram of", col),
         xlab = col, col = "blue", border = "red")
    qqnorm(dataframe[[col]], main = paste("QQ Plot of", col))
    qqline(dataframe[[col]])
  }
  par(mfrow = c(1, 1))
}

visnorm(df.clean.sub)
```

After omitting NAs, the data are not normally distributed. Nearly all variables exhibit skewness in one direction or another.

## Correlation Between Predictors

```{r eda-correlation}
corr <- round(cor(df.clean.sub[, -1]), 2)
corrplot(corr, method = "circle", type = "upper", order = "hclust",
         tl.col = "black", tl.srt = 45)
```

Key observations:

- **Productivity** and **P.gf.** are weakly correlated (~-0.11).
- **Inference_Counts** and **Inference_Sources** are moderately to strongly correlated.
- **Average_C10** and **Average_LogC10** are strongly correlated, as expected from their shared basis.

# Linear Regression Analysis

## Baseline Model (All Predictors)

```{r lm-baseline}
lm.h.index <- lm(H.index ~ ., data = df.clean.sub)
summary(lm.h.index)
```

**Productivity**, **Average_C10**, **Average_LogC10**, and **Inference_Sources** are statistically significant predictors of H-index. **Inference_Counts** and **P.gf.** are not significant in this model.

## Investigating Productivity and Gender

```{r lm-productivity-gender}
lm.prod.h.index <- lm(H.index ~ P.gf. + Productivity, data = df.clean.sub)
lm.prod.h.index.2 <- lm(H.index ~ Productivity, data = df.clean.sub)
lm.prod.h.index.3 <- lm(H.index ~ P.gf., data = df.clean.sub)

summary(lm.prod.h.index)
summary(lm.prod.h.index.2)
summary(lm.prod.h.index.3)
```

Due to the negative correlation between Productivity and P.gf., gender probability loses significance when both are included. Individually, both are significant predictors of H-index. However, the P.gf.-only model explains very little variance (low R-squared).

## Citation and Inference Model

```{r lm-citation-inference}
lm.h.index.c10.infer <- lm(
  H.index ~ Average_C10 + Average_LogC10 + Inference_Sources + Inference_Counts,
  data = df.clean.sub
)
summary(lm.h.index.c10.infer)
```

The predictors collectively have a significant effect, but the model explains only a modest amount of variability in H-index.

## Full Model with Interaction Terms

```{r lm-interaction-full}
lm.prod.full <- lm(H.index ~ .^2 - 1, data = df.clean.sub)
summary(lm.prod.full)
```

Many interaction terms are significant, and the model explains substantially more variance. Notably, while P.gf. alone has borderline significance, its interaction with Productivity is highly significant -- a surprising result given the earlier finding.

```{r lm-interaction-subset}
lm.prod.full.sub <- lm(
  H.index ~ Productivity + P.gf. + Productivity * P.gf. +
    Productivity^2 + P.gf.^2,
  data = df.clean.sub
)
summary(lm.prod.full.sub)
```

When including squared and interaction terms for just Productivity and P.gf., all outcomes become significant. This model explains nearly 80% of the variability in H-index.

## Quadratic Terms Only

```{r lm-quadratic}
lm.prod.full.quad <- lm(
  H.index ~ I(Productivity^2) + I(Average_C10^2) + I(Average_LogC10^2) +
    I(Inference_Sources^2) + I(Inference_Counts^2) + I(P.gf.^2),
  data = df.clean.sub
)
summary(lm.prod.full.quad)
```

The quadratic-only model explains ~53% of the variation, compared to ~93% for the interaction model. The quadratic term for Average_C10 is not significant here (p = 0.22), whereas it is highly significant in the interaction model.

# Model Diagnostics

## Variance Inflation Factors (VIF)

```{r diagnostics-vif}
print(vif(lm.h.index))
```

VIF values indicate little to moderate multicollinearity among predictors in the baseline model.

## Normality of Residuals

```{r diagnostics-normality}
print(shapiro.test(lm.h.index$residuals))
```

The Shapiro-Wilk test rejects the null hypothesis: residuals are **not** normally distributed.

## Constant Variance (Homoscedasticity)

```{r diagnostics-variance}
plot(lm.h.index$fitted.values, lm.h.index$residuals,
     xlab = "Fitted Values", ylab = "Residuals",
     main = "Residuals vs. Fitted Values")
print(bptest(lm.h.index))
```

The Breusch-Pagan test rejects constant variance. There is evidence of **heteroscedasticity** in the model, confirmed visually by the residual plot.

## Leverage Points

```{r diagnostics-leverage}
X <- model.matrix(lm.h.index)
H <- X %*% solve(t(X) %*% X) %*% t(X)
plot(diag(H), main = "Hat Values (Leverage)", ylab = "Leverage", xlab = "Observation")
abline(h = 2 * 7 / 5000, col = 1)
text(x = 1:length(diag(H)), y = diag(H),
     labels = 1:length(diag(H)), pos = 2)
```

Observation **3845** has the largest leverage.

## Influential Points (Cook's Distance)

```{r diagnostics-cooks}
cook.dist.prod <- cooks.distance(lm.h.index)
halfnorm(cook.dist.prod, 2, ylab = "Cook's Distance")
```

Observation **3845** is also the most influential point by Cook's distance.

## Outlier Detection

```{r diagnostics-outliers}
estar <- rstudent(lm.h.index)
plot(lm.h.index$fitted.values, estar,
     xlab = "Fitted Values", ylab = "Studentized Residuals",
     main = "Studentized Residuals vs. Fitted Values")
abline(h = 2 * 7 / 5000, col = 1)
text(x = lm.h.index$fitted.values, y = estar,
     labels = 1:length(lm.h.index$fitted.values), pos = 2)
```

Observation **3441** appears to be an outlier based on studentized residuals.

# Variable Selection

## Best Subset Selection

```{r subset-selection}
require(leaps)
subset <- regsubsets(H.index ~ ., data = df.clean.sub)
rs <- summary(subset)
print(rs$which)
```

### Adjusted R-Squared

```{r subset-adjr2}
print(which.max(rs$adjr2))
plot(2:7, rs$adjr2,
     xlab = "No. of Parameters", ylab = "Adjusted R-squared",
     main = "Adjusted R-squared by Number of Parameters")
```

Five variables explain up to 83.26% of variance; the sixth adds negligible improvement.

### Mallows' Cp

```{r subset-cp}
names <- colnames(rs$which)
plot(2:7, rs$cp,
     xlab = "No. of Parameters", ylab = "Cp Statistic",
     main = "Mallows' Cp by Number of Parameters")
abline(a = 0, b = 1, col = "red")
text(2:7, rs$cp, labels = names, pos = 4, col = "blue", cex = 0.5)
```

The Cp criterion also supports a 5-parameter model, maintaining a good balance between complexity and explanatory power.

## Stepwise Regression (AIC)

```{r stepwise-aic}
stepwise.lm.h.index <- stepAIC(lm.h.index)
summary(stepwise.lm.h.index)
```

The lowest AIC (11927.72) selects **Productivity**, **Average_C10**, **Average_LogC10**, and **Inference_Sources**. The full model's AIC (11927.75) is nearly identical, indicating the reduced model provides comparable fit with fewer variables.

# Regularization Methods

## Ridge Regression

```{r ridge-regression}
df.clean.sub.scaled <- data.frame(scale(df.clean.sub))
lm.h.index.ridge <- lm.ridge(H.index ~ .,
                               data = df.clean.sub.scaled,
                               lambda = seq(0, 4000, 0.01))

GCV.lambda <- which.min(lm.h.index.ridge$GCV)
print(lm.h.index.ridge$lambda[GCV.lambda])

matplot(lm.h.index.ridge$lambda, coef(lm.h.index.ridge),
        type = "l",
        xlab = expression(lambda),
        ylab = expression(hat(beta)),
        col = 1,
        main = "Ridge Coefficient Paths")
abline(v = lm.h.index.ridge$lambda[GCV.lambda], col = "red")
```

The optimal lambda is **1.8** (via GCV), indicating the model fits the data closely with minimal shrinkage.

```{r ridge-coefficients}
select(lm.h.index.ridge)
lm.h.index.ridge <- lm.ridge(H.index ~ ., lambda = 1.8, df.clean.sub.scaled)
round(coef(lm.h.index.ridge), 4)
```

Modified HKB estimator: 0.918; Modified L-W estimator: 0.804.

## Non-Negative Garrote

```{r nng}
scaled.df.clean.sub <- scale(as.matrix(df.clean.sub))
X <- scaled.df.clean.sub[, -1]
y <- scaled.df.clean.sub[, 1]

a.cv <- cv.nnGarrote(x = X, y = y)
outcome <- nnGarrote(x = X, y = y, lambda.nng = a.cv$optimal.lambda.nng)

names.garrote <- c("(Intercept)", colnames(X))
name.df <- data.frame(
  Parameter = names.garrote,
  Coefficient = c(outcome$betas[1, 1], outcome$betas[-1, 1])
)
print(name.df)
```

The Non-Negative Garrote shrinks **Inference_Counts** and **P.gf.** to zero, retaining the remaining predictors as having non-zero impact on H-index.

## Lasso Regression

### LARS Algorithm (Cp Criterion)

```{r lasso-lars}
y.2 <- scale(df.clean.sub$H.index)
x.2 <- as.matrix(scale(df.clean.sub[, 2:7]))

a.lasso <- lars(x.2, y.2)
summary(a.lasso)

names.lasso <- colnames(x.2)
plot(1:7, a.lasso$Cp,
     xlab = "Step", ylab = "Cp",
     main = "Lasso Path - Cp Criterion")
abline(0, 1)
text(2:7, a.lasso$Cp, labels = names.lasso, pos = 3, col = "blue", cex = 0.5)
```

The Lasso path demonstrates good model performance in predictor selection and complexity.

### Lasso via glmnet (Cross-Validated)

```{r lasso-glmnet}
a.lasso <- glmnet(x.2, y.2, family = "gaussian")
a.cv <- cv.glmnet(x.2, y.2, family = "gaussian", nfolds = 5000)
a.cv$lambda.min
round(coef(a.lasso, s = a.cv$lambda.min), 5)
```

The optimal lambda is ~0.003. All predictors except **P.gf.** are positively associated with H-index.

# Summary

| Method | Key Finding |
|--------|------------|
| Baseline OLS | Productivity, Average_C10, Average_LogC10, and Inference_Sources are significant |
| Interaction Model | Explains ~93% of variance; P.gf. x Productivity interaction is significant |
| Quadratic Model | Explains ~53% of variance; less effective than interactions |
| Stepwise (AIC) | Selects 4 predictors; nearly identical AIC to full model |
| Ridge | Optimal lambda = 1.8; minimal shrinkage needed |
| Non-Negative Garrote | Shrinks Inference_Counts and P.gf. to zero |
| Lasso | Retains most predictors; P.gf. has smallest coefficient |

Across methods, **Productivity** and **Average_C10/Average_LogC10** consistently emerge as the strongest predictors of H-index, while **Inference_Counts** and **P.gf.** contribute the least.
